\subsection*{Significance of the study}
\addcontentsline{toc}{subsection}{Significance of the Study}
The significance of this study can contribute to building upon the existing
knowledge base about diabetes by studying the correlation between sleep
disturbances and blood sugar levels. Understanding the relationship between the
two variables can provide important clues for improving blood sugar control in
diabetes patients (DM). Researchers will be able to determine whether sleep
disturbance is a major cause of diabetes in non-diabetic people, which could
lead to further research on this topic. Furthermore, improving sleep quality to
prevent diabetes in non-diabetic people can potentially be an excellent
intervention, which is another important benefit of this study. The following
are the beneficiaries of this study:

\textbf{Diabetes patients.} This study will benefit Type 1 and Type 2 diabetes
patients by providing insights on the relationship between sleep quality and
blood sugar level. Improving the sleep quality may further improve and reduce
the blood sugar level of said patients.

\textbf{Healthcare professionals and providers.} By understanding the
significance of having good sleep quality, healthcare professionals and
providers can educate diabetes patients about the importance of sleep and its
impact on blood sugar levels. Better care and support with consideration for the
potential effects of sleep quality can be provided to patients, leading to
better management of diabetes.

\textbf{The Field of Medicine.} The study will provide significant knowledge
about the major risk factors for diabetes. This can potentially provide new
insights on how to prevent diabetes for non-diabetic people, which will
significantly reduce the rate of diabetes growth.

\textbf{General Public.} Educating the general public about the importance of
sleep quality can help prevent diabetes in non-diabetic people. This knowledge
has the potential to improve overall public health and reduce the burden of
diabetes on individuals and healthcare systems. Furthermore, the findings of
this study can also inform the development and implementation of interventions
aimed at improving sleep quality in diabetic patients.

\textbf{Future Researchers.} Recommendations from this study can provide new
innovations for future research regarding this topic. New studies can
potentially explore and develop interventions utilizing this knowledge for
diabetes patients.
